\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2024-04-17 docs/papers/sto/so.tex
Draft minor
Update paper draft

% AI for science and cosmology
Rapid advances in artificial intelligence (AI) and more specifically
deep learning (DL) have provided powerful tools that are revolutionizing
scientific research across disciplines.
AI for science consists of not only the state-of-art deep neural
networks, but also more fundamental computational tools and hardwares
like automatic differentiation (AD) and GPUs.
One particular application is the numerical simulation of the Universe.

% a brief review of (differentiable) cosmological simulations
BORG, ELUCID, and BORG-PM \citep{BORG, ELUCID, BORG-PM}
\texttt{FastPM} and \texttt{FlowPM} \citep{FastPM, vmad, SeljakEtAl2017, FlowPM}

% introduce PM and its limitation on small scales
The particle-mesh (PM) method solves the Poisson equation by scattering
the particles onto a mesh, on which FFT is performed and the Laplacian
operator reduces to simple multiplication of wavenumbers in Fourier
space.
Although being computationally efficient, the PM method is mainly
limited by its resolution on small scales, i.e. scales around the mesh
cell size.
This is due to the loss of information during the scattering process of
continuously distributed particles onto the discrete mesh.
We use the cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} to
compute the mass fractions of a particle scattered to nearby grid
points.
Other schemes include the nearest grid point (NGP) and triangle shaped
cloud (TSC) etc.

@@

% introduction of this work
The differentiable cosmological simulation with adjoint method
\citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
\citep{Li2022b} lay the foundation of the gradient-based optimization
discussed in this work.
Within this fully differentiable framework, we sharpen the gravitational
force by embedding neural networks (NNs) in the Poisson equation and
train them for the spatial optimization (SO) of the simulation
resolution on small scales.
For better interpretability, we further fit the trained NNs using
symbolic regression (SR).

@@

Cosmological $N$-body simulations evolve dark matter particles under the
gravitational force.
The dynamics of the particles is governed by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $- \vnabla \varphi$
follows
%
\begin{equation}
\left[- \vnabla \varphi\right]_\vk = - i \vk \varphi(\vk) =
                  i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\label{force}
\end{equation}
%
which then gives the force field in configuration space after an inverse
Fourier transform.

In the PM simulation, the density field $\delta(\vx)$ is estimated by
scattering the particles onto the mesh, and then the force on each
particle is estimated by gathering the force field from the mesh.
Therefore, the accuracy of the gravitational force estimated for each
particle depends on the resolution of the mesh.
To improve the accuracy of the PM solver, we introduce SO force
sharpening by modifying the force equation \eqref{force} through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of
the wavevector components and all possible relevant features
$\vvartheta$.
The subscript $i\in\{1,2,3\}$ denotes the 3 principal axes of the
simulation box.

@@

For both the $f$ and $g$ spatial optimization NNs (SONNs), we adopt
multilayer perceptron (MLP, i.e. NN with fully connected layers)
structures to learn the nonlinear functions.
Both $f$ and $g$ networks have $XXX$ hidden layers with $XXX\times n_{\rm
input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
input layer, which is [...] and [...] for $f$ and $g$ respectively.
The rectified linear unit (ReLU) activation function is used for hidden
layers.
We further apply the exponential function to the output as a regulator
for positive output.
Being embedded in the differentiable simulation library \pmwd, the NNs
can be trained using gradient-based optimization methods as in usual
deep learning tasks.
The parameters of the NNs, which enter the force evaluation in every
time integration step, can be treated just the same as other
cosmological parameters.
The objective gradients are accumulated, based on which the parameters
are then optimized.

>>>>

% AI for science and cosmology
Rapid advances in artificial intelligence (AI) and more specifically
deep learning have provided powerful tools that are revolutionizing
scientific research across disciplines.
AI for science benefits from not only the state-of-art architectures of
deep neural networks (NNs), but also more fundamental computational
tools and hardwares like automatic differentiation (AD) and GPUs.
One of the very promising areas of AI in Cosmology is the numerical
simulation of the Universe.

% combination of differentiable simulations, NN, and SR
NNs are flexible and efficient universal function approximators
\citep{Hornik1989}, which can be used to learn physical processes
when embedded in numerical simulations.
NNs are typically trained with gradient based optimizers, which requires
the simulations to be differentiable.

% a brief review of (differentiable) cosmological simulations
% BORG, ELUCID, and BORG-PM \citep{BORG, ELUCID, BORG-PM}
% \texttt{FastPM} and \texttt{FlowPM} \citep{FastPM, vmad, SeljakEtAl2017, FlowPM}

% introduce PM and its limitation on small scales
The particle-mesh (PM) method is a fundamental technique for
cosmological $N$-body simulations.
PM solves the Poisson equation by scattering the particles onto an
auxiliary mesh, on which the fast Fourier transform (FFT) is performed
and the Laplacian operator reduces to a simple multiplication of
wavenumbers in Fourier space.
The cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} has been
widely used to compute the mass fractions of a particle at $\vx'$
scattered to a nearby mesh grid point at $\vx$
%
\begin{equation}
  W(\vx, \vx') = \prod_{i=1}^3
    \max \Bigl( 1 - \frac{|x_i-x'_i|}{l_c}, 0 \Bigr),
  \label{cic}
  \end{equation}
%
where $l_c$ is the mesh cell size.
The gravitational forces on the mesh points are then grathered to form
the interpolated force for each particle, with the same weights as in
the scattering process.
Other scattering schemes include the nearest grid point (NGP) and
triangle shaped cloud (TSC) etc.
Although being computationally efficient, the PM method is limited by
its resolution on small scales, i.e. scales around and below the mesh
cell size.
This is due to the loss of information during the scattering and
gathering process between the continuously distributed particles and the
discrete mesh.

@@

% introduction of this work
The differentiable cosmological simulation with adjoint method
\citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
\citep{Li2022b} lay the foundation of the gradient-based optimization
discussed in this work.
Within this fully differentiable framework, we sharpen the gravitational
force by embedding NNs in the Poisson equation and train them for the
spatial optimization (SO) of the simulation resolution on small scales.
We further interpret the trained NNs using symbolic regression (SR).

@@

Cosmological $N$-body simulations evolve dark matter particles under the
gravitational force.
The dynamics of the particles is governed by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $\vF = - \vnabla \varphi$
follows
%
\begin{equation}
\vF(\vk) = - i \vk \varphi(\vk) =
                  i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\label{force}
\end{equation}
%
which then gives the force field in configuration space after an inverse
Fourier transform.

In the PM simulation, the density field $\delta(\vx)$ is estimated by
scattering the particles onto the mesh with e.g. the CIC scheme in
\eqref{cic}.
Then the force on each particle is estimated by gathering the force
field from the mesh.
Therefore, the accuracy of the gravitational force estimated for each
particle depends on the resolution of the mesh.
To improve the accuracy of the PM solver, we introduce SO force
sharpening by modifying the force equation \eqref{force} to
%
\begin{equation}
F(\vk)_i = i \frac32 \Omegam H_0^2 \,\frac{k_i}{k^2}\,
                    f(k_i; \vvartheta) g(\vk; \vvartheta) \delta(\vk),
\end{equation}
%
where $f$ and $g$ are nonlinear functions of the wavevector components
and all possible relevant features $\vvartheta$, which are approximated
with NNs.
$k_i$ is a component of $\vk$, with $i\in\{1,2,3\}$ denoting the 3
principal axes of the simulation box.

@@

For both the $f$ and $g$ spatial optimization NNs (SONNs), the input
features consist of tabular data, therefore we adopt multilayer
perceptron (MLP, i.e. NN with fully connected layers) structures to
learn the nonlinear functions.
Both $f$ and $g$ networks have $XXX$ hidden layers with $XXX\times n_{\rm
input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
input layer, which is [...] and [...] for $f$ and $g$ respectively.
The rectified linear unit (ReLU) activation function is used for hidden
layers.
We further apply the exponential function to the output as a regulator
for positive output.
Being embedded in the differentiable simulation library \pmwd, the NNs
can be trained using gradient-based optimization methods as in usual
deep learning tasks.
The parameters of the NNs, which enter the force evaluation in every
time integration step, can be treated just the same as other
cosmological parameters.
The objective gradients are accumulated, based on which the parameters
are then optimized.
